---
title: "Comparison with other methods"
author: "Wei Jiang"
date: "5/10/2018"
output: html_notebook
---
We want to now assess the performance of algorithm SAEM and compare SAEM to several other existing methods for missing data.

```{r}
library(misaem)
library(MASS)
library(mvtnorm)
library(mice)
```

Here we first assign the true values of parameters.

(By using different values, we can construct different setting, such as number of subjects $n$ or structure of correlation.)
```{r}
n <- 1000  # number of subjects
# n <- 10000 # or a larger number of subjects 

p <- 5     # number of explanatory variables
mu.star <- 1:p  # mean of the explanatory variables
sd <- 1:p # standard deviations

# with correlation
C <- matrix(c(   # correlation matrix
  1,   0.8, 0,   0,   0,
  0.8, 1,   0,   0,   0,
  0,   0,   1,   0.3, 0.6,
  0,   0,   0.3, 1,   0.7,
  0,   0,   0.6, 0.7, 1
), nrow=p)
## or without correlation
# C = diag(p)

Sigma.star <- diag(sd)%*%C%*%diag(sd) # variance-covariance matrix of the explanatory variables

beta.star <- c(0.5, -0.3, 1, 0, -0.6) # coefficients of logistic regression
beta0.star <- -0.2  # intercept
beta.true = c(beta0.star,beta.star)

#percentage of missingness
p.miss <- 0.10 
patterns = runif(n*p)<p.miss

```

We considered the following competitors:
* The complete case (CC) method : all rows containing at least one unobserved data value were removed)
* Multiple imputation based on conditional modeling as implemented in the R package *mice* (with its default settings and Rubin's combining rules)
* MCEM algorithm that we implemented using adaptive rejection sampling (MCEM-AR). 
* We use the dataset without missing values (no NA) as a reference, with parameters estimated with the Newton-Raphson algorithm as implemented in the *glm* function in R.

We run repetitions of simulations. And evaluate their performance, intially in terms of estimation errors of the parameters, as well as the standard error of estimation and the coverage of confidence interval.
```{r}
nbsim = 1000
EST.saem = EST.comp = EST.cc = EST.mice = matrix(0, nbsim,length(beta.star)+1)
TIME.saem = TIME.mice = rep(0, nbsim)
STD.saem = STD.comp = STD.cc = STD.mice = matrix(0, nbsim,length(beta.star)+1)
LENGTH.saem = LENGTH.comp = LENGTH.cc = LENGTH.mice= matrix(0, nbsim,length(beta.star)+1)
count.saem = count.comp = count.cc =  count.mice = rep(0,p+1)

for (NB in 1:nbsim){
  set.seed(NB)
  # complete data simulation
  X.complete <- matrix(rnorm(n*p), nrow=n)%*%chol(Sigma.star) + matrix(rep(mu.star,n), nrow=n, byrow = TRUE)
  p1 <- 1/(1+exp(-X.complete%*%beta.star-beta0.star))
  y <- as.numeric(runif(n)<p1)
  
  # ----- No NA : classical estimation in the case without missingness
  data.complete <- data.frame(y=y,X.complete)
  model.complete <- glm(y ~.,family=binomial(link='logit'),data=data.complete)
  beta0.complete <- model.complete$coefficients[1]
  beta.complete <- model.complete$coefficients[2:(p+1)]
  P <- predict(model.complete, type = "response")
  W <- diag(P*(1-P))
  X <- model.matrix(model.complete)
  V_complete <- solve(t(X)%*%W%*%X)
  std.complete <- sqrt(diag(V_complete))
  
  # generating missing data - MCAR missingness
  X.obs <- X.complete
  X.obs[patterns] <- NA
  
  ## generating missing data - MAR missingness
  # X.obs <- X.complete
  # for(i in c(2,4,5)){
  #   z <- cbind(y,X.complete[,c(1,3)])%*%matrix(sample(-5:5, 3, replace=T),ncol=1) # linear combination 
  #   pr <- 1/(1+exp(-z))         # pass through an inv-logit function
  #   r <- rbinom(n,1,pr)      # bernoulli response variable
  #   X.obs[r==0,i]<-NA
  # }
  # cat('percentage of NA: ', mean(is.na(X.obs[,2])),mean(is.na(X.obs[,4])),mean(is.na(X.obs[,5])),'\n')
  # 
  
  # ------- CC : estimation ignoring the missing data
  data.obs <- data.frame(y=y,X.obs)
  model.obs <- glm(y ~.,family=binomial(link='logit'),data=data.obs)
  beta0.cc <- model.obs$coefficients[1]
  beta.cc <- model.obs$coefficients[2:(p+1)]
  P <- predict(model.obs, type = "response")
  W <- diag(P*(1-P))
  X <- model.matrix(model.obs)
  V_cc <- solve(t(X)%*%W%*%X)
  std.cc <- sqrt(diag(V_cc))
  
  # ------- mice : multiple imputation
  ptm <- Sys.time()
  DATA.ch= cbind.data.frame(y,X.obs)
  imp.ch <- mice(DATA.ch,print = FALSE)
  fit.ch <- glm.mids(y~., data=imp.ch,family = binomial)
  beta.mice=summary(pool(fit.ch, method = "rubin1987"))[,1]
  std.mice=summary(pool(fit.ch, method = "rubin1987"))[,2]
  time.mice=Sys.time() - ptm
  
  # -------  SAEM
  list.saem=miss.saem(X.obs,y, print_iter=FALSE,var_cal=TRUE)
  beta.saem = list.saem$beta
  std.saem = list.saem$std_obs
  
  EST.comp[NB,] = c(beta0.complete,beta.complete)
  EST.cc[NB,] = c(beta0.cc,beta.cc)
  EST.saem[NB,] = beta.saem
  EST.mice[NB,] = beta.mice
  
  STD.comp[NB,] = std.complete
  STD.cc[NB,] = std.cc
  STD.saem[NB,] = std.saem
  STD.mice[NB,] = std.mice
  
  TIME.saem[NB] = list.saem$time_run
  TIME.mice[NB] = time.mice
  
  ci.comp_ceil =  c(beta0.complete,beta.complete) + 1.96*std.complete
  ci.comp_ground =  c(beta0.complete,beta.complete) - 1.96*std.complete
  ci.cc_ceil =  c(beta0.cc,beta.cc) + 1.96*std.cc
  ci.cc_ground =  c(beta0.cc,beta.cc) - 1.96*std.cc
  ci.saem_ceil = beta.saem + 1.96*std.saem
  ci.saem_ground = beta.saem - 1.96*std.saem
  ci.mice_ceil = beta.mice + 1.96*std.mice
  ci.mice_ground = beta.mice - 1.96*std.mice
  
  LENGTH.comp[NB,] = ci.comp_ceil - ci.comp_ground
  LENGTH.cc[NB,] = ci.cc_ceil - ci.cc_ground
  LENGTH.saem[NB,] = ci.saem_ceil - ci.saem_ground
  LENGTH.mice[NB,] = ci.mice_ceil - ci.mice_ground
  for(i in 1:(p+1)){
    if( ci.comp_ground[i] <=beta.true[i] & ci.comp_ceil[i]>beta.true[i]){
      count.comp[i]<-count.comp[i]+1
    }
    if( ci.cc_ground[i] <=beta.true[i] & ci.cc_ceil[i]>beta.true[i]){
      count.cc[i]<-count.cc[i]+1
    }
    if( ci.saem_ground[i] <=beta.true[i] & ci.saem_ceil[i]>beta.true[i]){
      count.saem[i]<-count.saem[i]+1
    }
    if( ci.mice_ground[i] <=beta.true[i] & ci.mice_ceil[i]>beta.true[i]){
      count.mice[i]<-count.mice[i]+1
    }
  }
  cat('simulation =',NB,', count(complete) =',count.comp[2],', count(cc) =',count.cc[2],', count(mice) =',count.mice[2],', count(saem) =',count.saem[2], '\n')
}
```
We compare the bias and the standard error of estimation by observing the boxplots.
```{r}
par(mfrow=c(2,3))
boxplot(EST.comp[,1]-beta0.star,EST.cc[,1]-beta0.star,EST.mice[,1]-beta0.star,EST.saem[,1]-beta0.star,main='bias for beta0',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)
boxplot(EST.comp[,2]-beta.star[1],EST.cc[,2]-beta.star[1],EST.mice[,2]-beta.star[1],EST.saem[,2]-beta.star[1],main='bias for beta1',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)
boxplot(EST.comp[,3]-beta.star[2],EST.cc[,3]-beta.star[2],EST.mice[,3]-beta.star[2],EST.saem[,3]-beta.star[2],main='bias for beta2',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)
boxplot(EST.comp[,4]-beta.star[3],EST.cc[,4]-beta.star[3],EST.mice[,4]-beta.star[3],EST.saem[,4]-beta.star[3],main='bias for beta3',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)
boxplot(EST.comp[,5]-beta.star[4],EST.cc[,5]-beta.star[4],EST.mice[,5]-beta.star[4],EST.saem[,5]-beta.star[4],main='bias for beta4',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)
boxplot(EST.comp[,6]-beta.star[5],EST.cc[,6]-beta.star[5],EST.mice[,6]-beta.star[5],EST.saem[,6]-beta.star[5],main='bias for beta5',names=c("no NA","CC","mice","saem"))
abline(h = 0, col = "red", lty = 2)

par(mfrow=c(2,3))
boxplot(LENGTH.comp[,1],LENGTH.mice[,1],LENGTH.saem[,1],main='length of CI of beta0',names=c("no NA","mice",'SAEM'))
boxplot(LENGTH.comp[,2],LENGTH.mice[,2],LENGTH.saem[,2],main='length of CI of beta1',names=c("no NA","mice",'SAEM'))
boxplot(LENGTH.comp[,3],LENGTH.mice[,3],LENGTH.saem[,3],main='length of CI of beta2',names=c("no NA","mice",'SAEM'))
boxplot(LENGTH.comp[,4],LENGTH.mice[,4],LENGTH.saem[,4],main='length of CI of beta3',names=c("no NA","mice",'SAEM'))
boxplot(LENGTH.comp[,5],LENGTH.mice[,5],LENGTH.saem[,5],main='length of CI of beta4',names=c("no NA","mice",'SAEM'))
boxplot(LENGTH.comp[,6],LENGTH.mice[,6],LENGTH.saem[,6],main='length of CI of beta5',names=c("no NA","mice",'SAEM'))

par(mfrow=c(1,1))
boxplot(TIME.mice[1:25],TIME.saem[1:25],main='execution time',names=c('mice','SAEM'))

par(mfrow=c(2,3))
boxplot(STD.comp[,1],STD.mice[,1],STD.saem[,1],main='std for beta0',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,1])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,1])), col = "green", lty = 3)
abline(h = sqrt(var(EST.comp[,1])), col = "blue", lty = 4)
boxplot(STD.comp[,2],STD.mice[,2],STD.saem[,2],main='std for beta1',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,2])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,2])), col = "green", lty = 3)
abline(h = sqrt(var(EST.comp[,2])), col = "blue", lty = 4)
boxplot(STD.comp[,3],STD.mice[,3],STD.saem[,3],main='std for beta2',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,3])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,3])), col = "green", lty = 3)
abline(h = sqrt(var(EST.comp[,3])), col = "blue", lty = 4)
boxplot(STD.comp[,4],STD.mice[,4],STD.saem[,4],main='std for beta3',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,4])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,4])), col = "green", lty = 2)
abline(h = sqrt(var(EST.comp[,4])), col = "blue", lty = 2)
boxplot(STD.comp[,5],STD.mice[,5],STD.saem[,5],main='std for beta4',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,5])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,5])), col = "green", lty = 3)
abline(h = sqrt(var(EST.comp[,5])), col = "blue", lty = 4)
boxplot(STD.comp[,6],STD.mice[,6],STD.saem[,6],main='std for beta5',names=c("no NA","mice",'SAEM'))
abline(h = sqrt(var(EST.saem[,6])), col = "red", lty = 2)
abline(h = sqrt(var(EST.mice[,6])), col = "green", lty = 3)
abline(h = sqrt(var(EST.comp[,6])), col = "blue", lty = 4)
```

